---
format: 
  gfm:
    toc: true
    toc-depth: 3
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    message: false
    warning: false
    fig.path: "man/figures/README-"
---

<!-- README.md is generated from README.qmd. Please edit that file -->

# nflendzonePipeline

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**Automated NFL game predictions and team strength estimates using Bayesian state-space models.**

This package provides an automated pipeline for NFL analytics, with weekly updates powered by GitHub Actions. The visualizations below are automatically updated each week with the latest predictions and team rankings.

## Installation

You can install the development version of nflendzonePipeline from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("TylerPollard410/nflendzonePipeline")
```

# Weekly Report

This report provides weekly updates on NFL team strength estimates, home field advantage, and game predictions using Bayesian state-space models.

```{r}
#| label: libraries

library(tidyverse)
library(posterior)
library(distributional)
library(tidybayes)
library(ggdist)
library(bayesplot)
library(patchwork)
library(nflreadr)
library(nflplotR)
library(nflendzone)
library(nflendzoneModel)
library(nflendzonePipeline)

theme_set(theme_ggdist())
```

## Data Setup

Load game data and team information from all available seasons.

```{r}
#| label: globals
#| code-fold: true
#| code-summary: "Show the R code - globals"

all_seasons <- 2002:nflreadr::most_recent_season()
teams_data <- nflreadr::load_teams(current = TRUE)
game_data <- load_game_data(seasons = all_seasons)
game_data_long <- load_game_data_long(game_df = game_data)
season_weeks_df <- game_data |>
  dplyr::distinct(season, week, week_seq)

base_repo_url <-
  "https://github.com/TylerPollard410/nflendzoneData/releases/download/"
```

## Load Model Estimates

Extract the latest filtered and predicted estimates from the data repository.

```{r}
#| label: load-estimates-data-function
#| code-fold: true
#| code-summary: "Show the R code - load-estimates-data-function"

# Function to load timestamps and estimates for a given set of tags
load_estimates_data <- function(tags, base_url) {
  # Load timestamps
  timestamps <- purrr::map(
    tags,
    ~ {
      timestamp_url <-
        paste0(base_url, .x, "/", .x, "_timestamp.json")
      jsonlite::fromJSON(timestamp_url)
    }
  ) |>
    purrr::set_names(tags)

  # Load estimates using season and week from timestamp data
  estimates <- purrr::imap(
    timestamps,
    ~ {
      # Extract season and week from timestamp
      season <- .x$season
      week <- .x$week
      week_idx <- .x$week_idx

      # Build URL with season and week in filename
      data_url <-
        paste0(base_url, .y, "/", .y, "_", season, "_", week, ".rds")
      data <- nflreadr::rds_from_url(data_url)

      # Attach timestamp as attributes
      attr(data, "season") <- season
      attr(data, "week") <- week
      attr(data, "week_idx") <- week_idx

      return(data)
    }
  )

  return(estimates)
}
```

```{r}
#| label: extract-estimates
#| code-fold: true
#| code-summary: "Show the R code - extract-estimates"

# Define both sets of tags
filter_tags <- c(
  "team_strength_filter",
  "league_hfa_filter",
  "result_filter"
)

predict_tags <- c(
  "team_strength_predict",
  "league_hfa_predict",
  "result_predict"
)

# Load filter data
filter_data <- load_estimates_data(filter_tags, base_repo_url)

# Load predict data
predict_data <- load_estimates_data(predict_tags, base_repo_url)
```

```{r}
#| label: clean-data
#| code-fold: true
#| code-summary: "Show the R code - clean-data"

# Clean up filter data
filter_data <- filter_data |>
  purrr::map(
    \(x) {
      x |>
        dplyr::rename_with(
          ~ stringr::str_remove(.x, "^filtered_")
        ) |>
        mutate(type = "filter")
    }
  )


# Clean up predict data
predict_data <- predict_data |>
  purrr::map(
    \(x) {
      x |>
        dplyr::rename_with(
          ~ stringr::str_remove(.x, "^predicted_")
        ) |>
        mutate(type = "predict")
    }
  )
```

## Team Strength Rankings

Current team strength estimates ranked from strongest to weakest. Values
represent the expected point differential against an average team on a neutral
field. The gradient intervals show the full posterior distribution.

```{r}
#| label: team-strength-plot
#| code-fold: true
#| code-summary: "Show the R code - team-strength-plot"
#| fig-align: center
#| fig-width: 10
#| fig-height: 14

# Create named vector of team colors (lightened for visibility)
team_colors <- setNames(teams_data$team_color, teams_data$team_abbr)
team_colors_light <- colorspace::lighten(team_colors, amount = 0.25)

# Filter for the latest week
team_strength_filter_plot <- filter_data |>
  pluck("team_strength_filter") |>
  left_join(teams_data, by = c("team" = "team_abbr")) |>
  arrange(desc(E(team_strength))) |>
  mutate(rank = row_number()) |>
  ggplot(aes(y = reorder(team, team_strength))) +
  # Zero reference line (average team)
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1
  ) +
  # Gradient interval showing uncertainty - lightened colors for better visibility
  stat_gradientinterval(
    aes(xdist = team_strength, fill = team),
    scale = 0.8,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = team_colors_light) +
  # Team logos
  geom_nfl_logos(
    aes(x = -14, team_abbr = team),
    width = 0.045
  ) +
  # Rank numbers
  geom_text(
    aes(x = -17, label = rank),
    size = 5,
    fontface = "bold",
    color = "gray10"
  ) +
  # Expected value labels - positioned to the right, outside the gradient
  geom_label(
    aes(x = 16, label = sprintf("%.1f", E(team_strength))),
    size = 4.5,
    fontface = "bold",
    label.size = 0,
    fill = "white",
    alpha = 0.9,
    label.padding = unit(0.2, "lines"),
    position = position_jitter(width = 0, height = 0.15, seed = 42)
  ) +
  scale_x_continuous(
    breaks = seq(-15, 15, 5),
    limits = c(-18, 18)
  ) +
  labs(
    title = "NFL Team Strength Rankings",
    subtitle = paste(
      "Season",
      attr(filter_data$team_strength_filter, "season"),
      "- Week",
      attr(filter_data$team_strength_filter, "week"),
      "| Expected point differential vs average team on neutral field"
    ),
    x = "Team Strength (Points)",
    y = NULL,
    caption = "Gradient shows full posterior uncertainty | 0 = league average"
  ) +
  theme_ggdist() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5, color = "gray40"),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray90", linewidth = 0.3)
  )

team_strength_filter_plot
```

## Home Field Advantage Comparison

Team-specific home field advantages compared to league average. Values are
typically small (within ±2 points) showing modest variation around the league norm.

```{r}
#| label: hfa-comparison-plot
#| code-fold: true
#| code-summary: "Show the R code - hfa-comparison-plot"
#| fig-align: center
#| fig-width: 10
#| fig-height: 14

# Extract league HFA
league_hfa_filter <- filter_data |>
  pluck("league_hfa_filter")

# Combine team and league HFA for comparison
team_hfa_filter <- filter_data |>
  pluck("team_strength_filter")

# Create named vector of team colors
team_colors <- setNames(teams_data$team_color, teams_data$team_abbr)

hfa_comp_plot <- team_hfa_filter |>
  left_join(league_hfa_filter) |>
  mutate(hfa_diff = team_hfa - league_hfa) |>
  left_join(teams_data, by = c("team" = "team_abbr")) |>
  arrange(desc(E(hfa_diff))) |>
  mutate(rank = row_number()) |>
  ggplot(aes(y = reorder(team, hfa_diff))) +
  # Zero reference (league average)
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "gray30",
    linewidth = 1
  ) +
  # Uncertainty intervals - lightened team colors with dark intervals
  stat_pointinterval(
    aes(xdist = hfa_diff, fill = team),
    .width = c(0.5, 0.95),
    point_size = 2.5,
    interval_color = "gray20",
    color = "gray20",
    linewidth = 1.2,
    show.legend = FALSE
  ) +
  scale_fill_manual(values = team_colors_light) +
  # Team logos
  geom_nfl_logos(
    aes(x = -3.5, team_abbr = team),
    width = 0.045
  ) +
  # Rank numbers
  geom_text(
    aes(x = -4.3, label = rank),
    size = 5,
    fontface = "bold",
    color = "gray10"
  ) +
  # Expected value labels - positioned to the right, outside the intervals
  geom_label(
    aes(x = 4.2, label = sprintf("%+.1f", E(hfa_diff))),
    size = 4,
    fontface = "bold",
    label.size = 0,
    fill = "white",
    alpha = 0.9,
    label.padding = unit(0.2, "lines"),
    position = position_jitter(width = 0, height = 0.15, seed = 42)
  ) +
  scale_x_continuous(
    breaks = seq(-4, 4, 1),
    limits = c(-4.5, 4.5)
  ) +
  labs(
    title = "Home Field Advantage Relative to League Average",
    subtitle = paste(
      "Season",
      attr(filter_data$team_strength_filter, "season"),
      "- Week",
      attr(filter_data$team_strength_filter, "week"),
      "| Difference from league HFA in points"
    ),
    x = "HFA Difference from League Average (Points)",
    y = NULL,
    caption = "Positive = stronger home advantage | Negative = weaker home advantage | 0 = league average"
  ) +
  theme_ggdist() +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5, color = "gray40"),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray90", linewidth = 0.3)
  )

hfa_comp_plot
```

## Weekly Game Predictions

Predicted outcomes for upcoming games with full uncertainty quantification.

```{r}
#| label: game-prediction
#| code-fold: true
#| code-summary: "Show the R code - game-prediction"

result_predict <- predict_data |>
  pluck("result_predict")
# mutate(
#   y2 = rvar_rng(rnorm, n = n(), mean = mu, sd = sigma, ndraws = 10000),
#   y3 = mu + rvar_rng(rnorm, n = n(), mean = 0, sd = sigma),
#   y4 = mu + rvar_rng(rnorm, n = 1, mean = 0, sd = 1) * sigma,
#   .after = y
# )

pred_df <- inner_join(
  result_predict,
  game_data
) |>
  mutate(
    home_mu_cover_prob = Pr(mu > spread_line),
    home_y_cover_prob = Pr(y > spread_line),
    away_mu_cover_prob = Pr(mu < spread_line),
    away_y_cover_prob = Pr(y < spread_line),
    mu_cover_prob = pmax(home_mu_cover_prob, away_mu_cover_prob),
    y_cover_prob = pmax(home_y_cover_prob, away_y_cover_prob),
    mu_cover_team = case_when(
      home_mu_cover_prob > away_mu_cover_prob ~ home_team,
      home_mu_cover_prob < away_mu_cover_prob ~ away_team,
      TRUE ~ NA_character_
    ),
    y_cover_team = case_when(
      home_y_cover_prob > away_y_cover_prob ~ home_team,
      home_y_cover_prob < away_y_cover_prob ~ away_team,
      TRUE ~ NA_character_
    ),
    mu_bet_team = ifelse(mu_cover_prob > 0.55, mu_cover_team, NA_character_),
    y_bet_team = ifelse(y_cover_prob > 0.55, y_cover_team, NA_character_)
  )
```

```{r}
#| label: prep-prediction-data
#| code-fold: true
#| code-summary: "Show the R code - prep-prediction-data"

# Create named vector of team colors
team_colors <- setNames(teams_data$team_color, teams_data$team_abbr)
team_colors2 <- setNames(teams_data$team_color2, teams_data$team_abbr)

# Prepare data with better formatting
pred_plot_df <- pred_df |>
  mutate(
    matchup = paste0(away_team, " @ ", home_team),
    matchup_display = if_else(
      hfa == 1,
      paste0(away_team, " @ ", home_team),
      paste0(away_team, " vs ", home_team, " (N)")
    )
  ) |>
  arrange(game_idx)
```

### Expected Point Spreads vs Betting Lines

How our model's expected spread compares to Vegas betting lines. Points show the
expected differential with uncertainty intervals.

```{r}
#| label: spread-comparison-plot
#| code-fold: true
#| code-summary: "Show the R code - spread-comparison-plot"
#| fig-align: center
#| fig-width: 11
#| fig-height: 8

spread_plot <- pred_plot_df |>
  ggplot(aes(y = reorder(matchup_display, game_idx))) +
  # Zero reference line
  geom_vline(xintercept = 0, linetype = "solid", color = "gray60", linewidth = 0.5) +
  # Predicted game outcome (y) - wider uncertainty including game randomness
  stat_pointinterval(
    aes(xdist = y),
    .width = c(0.5, 0.95),
    point_size = 3.5,
    linewidth = 1.3,
    color = "gray50",
    alpha = 0.4
  ) +
  # Expected value (mu) - tighter uncertainty, parameter only
  stat_pointinterval(
    aes(xdist = mu),
    .width = c(0.66, 0.95),
    point_size = 3.5,
    linewidth = 1.3,
    color = "#013369"
  ) +
  # Betting spread line - on top so visible
  geom_point(
    aes(x = spread_line),
    color = "#D50A0A",
    size = 5.5,
    shape = 18
  ) +
  # Add team logos
  geom_nfl_logos(
    aes(x = -32, team_abbr = away_team),
    width = 0.04
  ) +
  geom_nfl_logos(
    aes(x = 32, team_abbr = home_team),
    width = 0.04
  ) +
  scale_x_continuous(
    breaks = seq(-30, 30, 10),
    limits = c(-35, 35)
  ) +
  labs(
    title = "Model Predictions vs Betting Lines",
    subtitle = paste(
      "Season",
      attr(predict_data$result_predict, "season"),
      "- Week",
      attr(predict_data$result_predict, "week"),
      "| Gray: Full prediction (y) | Blue: Expected value (mu) | Red: Vegas line"
    ),
    x = "Point Differential (Positive = Home Team Favored)",
    y = NULL,
    caption = "Gray shows game prediction uncertainty | Blue shows model's best estimate | Use y (gray) for betting decisions"
  ) +
  theme_ggdist() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5, color = "gray40"),
    axis.text.y = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank()
  )

spread_plot
```

### Win Probability by Game

Probability that the home team wins each matchup.

```{r}
#| label: win-prob-plot
#| code-fold: true
#| code-summary: "Show the R code - win-prob-plot"
#| fig-align: center
#| fig-width: 11
#| fig-height: 8

win_prob_plot <- pred_plot_df |>
  mutate(
    home_win_prob = Pr(y > 0),
    away_win_prob = 1 - home_win_prob,
    favored_team = if_else(home_win_prob > 0.5, home_team, away_team),
    favored_color = if_else(home_win_prob > 0.5, team_colors[home_team], team_colors[away_team])
  ) |>
  ggplot(aes(y = reorder(matchup_display, game_idx))) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_col(
    aes(x = home_win_prob, fill = favored_team),
    width = 0.75
  ) +
  geom_text(
    aes(
      x = home_win_prob,
      label = scales::percent(home_win_prob, accuracy = 1)
    ),
    hjust = -0.2,
    size = 4,
    fontface = "bold"
  ) +
  geom_nfl_logos(
    aes(x = 0.02, team_abbr = away_team),
    width = 0.035
  ) +
  geom_nfl_logos(
    aes(x = 0.98, team_abbr = home_team),
    width = 0.035
  ) +
  scale_fill_manual(values = team_colors) +
  scale_x_continuous(
    labels = scales::percent,
    limits = c(0, 1.15),
    breaks = seq(0, 1, 0.25)
  ) +
  labs(
    title = "Home Team Win Probability",
    subtitle = paste(
      "Season",
      attr(predict_data$result_predict, "season"),
      "- Week",
      attr(predict_data$result_predict, "week")
    ),
    x = "Probability",
    y = NULL
  ) +
  theme_ggdist() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.text.y = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  )

win_prob_plot
```

### Predicted Score Distributions

Full predictive distribution for each game showing all possible outcomes.

```{r}
#| label: score-dist-plot
#| code-fold: true
#| code-summary: "Show the R code - score-dist-plot"
#| fig-align: center
#| fig-width: 11
#| fig-height: 9

score_dist_plot <- pred_plot_df |>
  ggplot(aes(y = reorder(matchup_display, game_idx))) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.6) +
  # Use y for full predictive distribution
  stat_halfeye(
    aes(xdist = y, fill = home_team),
    .width = c(0.5, 0.95),
    point_interval = "median_qi",
    alpha = 0.8,
    slab_linewidth = 0,
    interval_color = "gray20",
    point_color = "gray20",
    point_size = 2.5,
    linewidth = 1.5
  ) +
  geom_nfl_logos(
    aes(x = -40, team_abbr = away_team),
    width = 0.045
  ) +
  geom_nfl_logos(
    aes(x = 40, team_abbr = home_team),
    width = 0.045
  ) +
  scale_fill_manual(values = colorspace::lighten(team_colors, amount = 0.3)) +
  scale_x_continuous(
    breaks = seq(-40, 40, 10),
    limits = c(-45, 45)
  ) +
  labs(
    title = "Predicted Score Distributions (Game Outcomes)",
    subtitle = paste(
      "Season",
      attr(predict_data$result_predict, "season"),
      "- Week",
      attr(predict_data$result_predict, "week"),
      "| Full predictive distributions showing 50% and 95% credible intervals"
    ),
    x = "Point Differential (Positive = Home Team Wins)",
    y = NULL,
    caption = "Uses y (full prediction) including game randomness"
  ) +
  theme_ggdist() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5, color = "gray40"),
    axis.text.y = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    legend.position = "none"
  )

score_dist_plot
```

### Betting Opportunities

Games where our model disagrees with Vegas by at least 2 points or shows high confidence.

```{r}
#| label: betting-edge-plot
#| code-fold: true
#| code-summary: "Show the R code - betting-edge-plot"
#| fig-align: center
#| fig-width: 11
#| fig-height: 7

betting_plot <- pred_plot_df |>
  mutate(
    model_spread = E(y),  # Use y for betting decisions
    spread_diff = model_spread - spread_line,
    abs_diff = abs(spread_diff),
    bet_worthy = abs_diff >= 2 | y_cover_prob >= 0.60  # Use y_cover_prob
  ) |>
  filter(bet_worthy) |>
  ggplot(aes(y = reorder(matchup_display, abs_diff))) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray50", linewidth = 0.6) +
  # Arrow showing edge direction - draw first so it's under points
  geom_segment(
    aes(x = spread_line, xend = model_spread, yend = matchup_display),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
    linewidth = 1.8,
    color = "#013369",
    alpha = 0.7
  ) +
  # Vegas line - larger and on top
  geom_point(
    aes(x = spread_line),
    size = 6.5,
    color = "#D50A0A",
    shape = 18
  ) +
  # Model prediction - larger and on top
  geom_point(
    aes(x = model_spread),
    size = 5,
    color = "#013369"
  ) +
  # Edge label
  geom_text(
    aes(
      x = (spread_line + model_spread) / 2,
      label = sprintf("%.1f pts", abs(spread_diff))
    ),
    vjust = -0.9,
    size = 4.5,
    fontface = "bold",
    color = "gray20"
  ) +
  # Cover probability label
  geom_text(
    aes(
      x = (spread_line + model_spread) / 2,
      label = sprintf("%d%%", round(y_cover_prob * 100))
    ),
    vjust = 1.8,
    size = 3.5,
    fontface = "bold",
    color = "#013369"
  ) +
  geom_nfl_logos(
    aes(x = min(c(spread_line, model_spread)) - 5, team_abbr = away_team),
    width = 0.045
  ) +
  geom_nfl_logos(
    aes(x = max(c(spread_line, model_spread)) + 5, team_abbr = home_team),
    width = 0.045
  ) +
  labs(
    title = "Betting Edges Over Vegas Lines",
    subtitle = paste(
      "Season",
      attr(predict_data$result_predict, "season"),
      "- Week",
      attr(predict_data$result_predict, "week"),
      "| Games with 2+ point disagreement or 60%+ cover probability",
      "\nRed = Vegas | Blue = Model (y) | Percentages show cover probability"
    ),
    x = "Point Spread (Positive = Home Team Favored)",
    y = NULL,
    caption = "Uses y (full prediction) for realistic betting expectations"
  ) +
  theme_ggdist() +
  theme(
    plot.title = element_text(size = 17, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5, color = "gray40"),
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank()
  )

betting_plot
```

---

*This README is automatically updated via GitHub Actions each week with the latest predictions and team strength estimates.*
